The promise of community-driven preprints in ecology and evolution

### 2024-05-14 ms/ms.qmd
Added new ref and dealing with soem additonal comments

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While preprint use in biology has had a slower uptake [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings which can lead to more citations [e.g., @colavizza2024analysis]; 2) helping establish precedence of research findings; 3) removing financial barriers to publication; and 4) encouraging constructive and critical feedback from the research community [@Proulx2013; @bourne2017ten; @vale2015accelerating]. Preprints are expected to promote open science, increase the visibility of research (e.g., [@fu2019releasing]), and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally.

Preprint servers can empower researchers to make their research findings more accessible, open, and transparent but only if they are used as forums for spreading and discussing findings within a research community. Significant barriers to preprint use still exist globally – from a lack of clarity around preprint policies at journals [@Klebel2020], to a stigma that preprints are viewed by the research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution. Such an understanding can help inform future initiatives and allow us to work harder in further breaking down barriers to scientific publishing.

@@

Here, we explore the first 1000+ preprints uploaded to *EcoEvoRxiv* to understand preprint practices and uses in ecology and evolution. We focus our attention on better understanding: 1) in what countries authors who use *EcoEvoRxiv* are located; 2) how career stage and gender are associated with preprint server use; 3) the extent to which authors make use of preprint servers for registered reports and community-driven peer review; 4) the extent to which data and code are shared in preprints; and 5) how many preprints remain unpublished, and for those that are published, how long it took them to become published. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other preprint servers to help further educate researchers about the benefits of using community-driven preprint servers to post their research findings.

@@

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolutionary biology (SORTEE)[@o2021towards]. Originally launched in 2018 on the Center for Open Science preprint platform, *EcoEvoRxiv* has grown to become a popular preprint server for researchers in ecology and evolutionary biology. It has since been adopted by the California Digital Library (CDL) to be hosted in perpetuity without fees. Editors are researchers from across the globe (currently eight different countries) in the fields of ecology and evolutionary biology who volunteer their time to screen papers and push new initiatives in the preprint space. *EcoEvoRxiv* allows authors to post both preprints and postprints, or author-accepted manuscripts -  a version of an accepted or published article without typesetting and formatting by a journal. Postprints are a useful way to ensure published articles are openly accessible and can be posted anytime after publication (assuming the journal has a policy that recognises pre- and post-prints -- though most do; see https://www.sherpa.ac.uk/romeo/). To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the articles available on *EcoEvoRxiv* as of `r gsub("UTC", "", dates[2])` (see Supplement for more details on methods). For the purpose of summarising articles on *EcoEvoRxiv* we include both preprints and postprints as 'articles' in our overview analysis.

@@

After removing five duplicate titles – suggesting a few authors created multiple submissions for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55--60 preprints published bi-monthly in the last two years ([@fig-summary]A).  *EcoEvoRxiv* hosts papers from authors based in `r length(unique(countries$country))-1` countries, with 90% of the articles coming from just `r nrow(most_use)-1` countries. North America, Australia, and European countries upload the most preprints with much fewer coming from countries in Africa, Central America, and parts of Asia ([@fig-summary]B).  Articles covered all major taxonomic groups, with the most common being vertebrates (`r sum_taxa$perc[6]`%), plants (`r sum_taxa$perc[1]`%), and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]C).

>>>>

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While the adoption of preprints in biology have been slower relative to other disciplines [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed and visibility of research findings which can lead to more citations [e.g., @colavizza2024analysis]; 2) helping establish precedence of research findings; 3) removing financial barriers to open access publication; and 4) enabling feedback from the research community [@Proulx2013; @bourne2017ten; @vale2015accelerating]. Preprints are expected to promote open science, increase the visibility of research (e.g., [@fu2019releasing]), and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally [@ni2024preprint].

Preprint servers can empower researchers to make their research findings more accessible, open, and transparent but only if they are used as forums for spreading and discussing findings within a research community. Significant barriers to preprint use still exist globally, from a lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by the research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution. Such an understanding can help inform future initiatives and allow us to work harder in further breaking down barriers to scientific publishing.

@@

Here, we explore the first 1000+ preprints uploaded to *EcoEvoRxiv* to understand preprint practices and uses in ecology and evolution. We aim to better understand: 1) in what countries authors who use *EcoEvoRxiv* are located; 2) how career stage and gender are associated with preprint server use; 3) the extent to which authors make use of preprint servers for registered reports and community-driven peer review; 4) the extent to which data and code are shared in preprints; and 5) how many preprints remain unpublished, and for those that are published, how long it took them to become published. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other preprint servers to help further educate researchers about the benefits of using community-driven preprint servers to post their research findings.

@@

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolutionary biology (SORTEE)[@o2021towards]. Originally launched in 2018 on the Center for Open Science preprint platform, *EcoEvoRxiv* has grown to become a popular preprint server for researchers in ecology and evolutionary biology. The server has since been adopted by the California Digital Library (CDL) to be hosted in perpetuity without fees. Editors are researchers from across the globe (currently eight different countries) in the fields of ecology and evolutionary biology who volunteer their time to screen papers and push new initiatives in the preprint space. *EcoEvoRxiv* allows authors to post both preprints and postprints, or author-accepted manuscripts -  a version of an accepted or published article without typesetting and formatting by a journal. Postprints are a useful way to ensure published articles are openly accessible and can be posted anytime after publication (assuming the journal has a policy that recognises pre- and post-prints -- though most do; see https://www.sherpa.ac.uk/romeo/).

@@

To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the articles available on *EcoEvoRxiv* as of `r gsub("UTC", "", dates[2])` (see Supplement for more details on methods). For the purpose of summarising articles on *EcoEvoRxiv* we include both preprints and postprints as 'articles' in our overview analysis. After removing five duplicate titles – suggesting a few authors created multiple submissions for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55--60 preprints published bi-monthly in the last two years ([@fig-summary]A).  *EcoEvoRxiv* hosts papers from authors based in `r length(unique(countries$country))-1` countries, with 90% of the articles coming from just `r nrow(most_use)-1` countries. North America, Australia, and European countries upload the most preprints with much fewer coming from countries in Africa, Central America, and parts of Asia ([@fig-summary]B).  Articles covered all major taxonomic groups, with the most common being vertebrates (`r sum_taxa$perc[6]`%), plants (`r sum_taxa$perc[1]`%), and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]C).
